---
title: "Dplyr04_Summarizing_and_sl"
output: 
  md_document:
    toc: true
    toc_depth: 3

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, error = FALSE, warning=FALSE)
```


# Data Wrangling Part 4: Summarizing and slicing your data 

This is the fourth blog post in a series of dplyr tutorials:

* [Part 1: Basic to Advanced Ways to Select Columns](https://suzanbaert.netlify.com/2018/01/dplyr-tutorial-1/)
* [Part 2: Transforming your columns into the right shape](https://suzan.rbind.io/2018/02/dplyr-tutorial-2/)
* [Part 3: Filtering rows](https://suzan.rbind.io/2018/02/dplyr-tutorial-3/))


Content:  

--


Note: as per previous blog posts, I will present everything in the form of a pipe. In some of the below cases, this might not be necessary and it would be just as easy to write it as a single function, but as I want to present options that you can use in your pipes, all below examples will be piped.

<br>

**The data**  
As per previous blog posts many of these functions truly shine when you have a lot of columns, but to make it easy on people to copy paste code and experiment, I'm using a built-in dataset:
```{r}
library(dplyr)
msleep <- ggplot2::msleep
glimpse(msleep)
```
<br>

## **Counting cases and adding counts**

### Counting the number of observations


The easiest way to know how many observations you have for a specific variable, is to use `count()`. By adding the `sort = TRUE` argument, it immediately returns a sorted table with descending number of observations:

```{r}
msleep %>% 
  count(order, sort = TRUE)
```


You can add multiple variables to a `count()` statement; the example below is counting by order and vore:

```{r}
msleep %>% 
  count(order, vore, sort = TRUE)
```


### Tally, add_tally and add_count

If you're only interested in counting the total number of cases for a dataframe, you could use `tally()`, which behaves simarly to `nrow()`. 

You can't provide a variable to count with `tally()`, it only works to count the overall number of observations. In fact, as is described in the `dplyr` documentation, `count()` is a short-hand for `group_by()` and `tally()`. 

```{r}
msleep %>% 
  tally()
```

More interesting is the `add_tally()` function which automatically adds a column with the overall number of observations. This would be the same as `mutate(n = n())`.


```{r}
msleep %>% 
  select(1:3) %>% 
  add_tally()
```


Even more interesting is `add_count()` which takes a variable as argument, and adds a column which the number of observations. This saves the combination of grouping, mutating and ungrouping again.


```{r}
msleep %>% 
  select(name:vore) %>% 
  add_count(vore)
```





<br><hr>


## **Summarising data**


*To note: for some functions, `dplyr` foresees both an American English and a UK English variant. The function `summarise()` is the equivalent of `summarize()`.*

If you just want to know the number of observations `count()` does the job, but to produce summaries of the average, sum, standard deviation, minimum, maximum of the data, we need `summarise()`. To use the function you just add your new column name, and after the equal sign the mathematics of what needs to happen: `column_name = function(variable)`.
You can add multiple summary functions behind each other. 

```{r}
msleep %>% 
  summarise(n = n(), average = mean(sleep_total), maximum = max(sleep_total))
```


In most cases, we don't just want to summarise the whole data table, but we want to get summaries by a group. To do this, you first need to specify by which variable(s) you want to divide the data using `group_by()`. You can add one of more variables as arguments in `group_by()`. 

```{r}
msleep %>% 
  group_by(vore) %>% 
  summarise(n = n(), average = mean(sleep_total), maximum = max(sleep_total))
  
```



The `summarise()` call works with nearly any aggregate function, and allows for additional arithmetics:

+ `n()` - gives the number of observations
+ `n_distinct(var)` - gives the numbers of unique values of `var`
+ `sum(var)`, `max(var)`, `min(var)`, ...
+ `mean(var)`, `median(var)`, `sd(var)`, `IQR(var)`, ... 
+ ...

The sample code will average the sleep_total and divide by 24, to get the amount of sleep as a fraction of a day.


```{r}
msleep %>% 
  group_by(vore) %>% 
  summarise(avg_sleep_day = mean(sleep_total)/24)
```

<br><hr>


## **Summarise_all()**


Similarly to the filter, select and mutate functions, `summarise()` comes with three additional functions for doing things to multiple columns in one go:

+ `summarise_all()` will summarise all columns based on your further instructions
+ `summarise_if()` requires a function that returns a boolean. If that is true, the summary instructions will be followed
+ `sumarise_at()` requires you to specify columns inside a `vars()` argument for which the summary will be done.



### **Summarise_all**


The function `summarise_all()` requires a function as argument, which it will apply to all columns. The sample code first selects all numeric columns, and then calculates the mean for each of them. I had to add the `na.rm = TRUE` argument to ignore `NA` values. 

```{r}
msleep %>% 
  group_by(vore) %>% 
  summarise_all(mean, na.rm=TRUE) 
```


The instructions for summarizing have to be a function. When there is no funcion availble in base R or a package to do what you want, you can either make a function upfront, or make a function on the fly. 
The sample code will add 5 to the mean of each column. The function on the fly can be made by either using `funs(mean(., na.rm = TRUE) + 5)`, or via a tilde: `~mean(., na.rm = TRUE) + 5`.



```{r}
msleep %>%
  group_by(vore) %>% 
  summarise_all(~mean(., na.rm = TRUE) + 5) 
```
<br>


### **Summarise_if**

The function `summarise_if()` requires two arguments: 

+ First it needs information about the columns you want it to consider. This information needs to be a function that returns a boolean value. The easiest cases are functions like `is.numeric`, `is.integer`, `is.double`, `is.logical`, `is.factor`, `lubridate::is.POSIXt` or `lubridate::is.Date`.

+ Secondly, it needs information about how to summarise that data, which as above needs to be a function. If not a function, you can create a function on the fly using `funs()` or a tilde (see above).


The sample code below will return the average of all numeric columns:


```{r}
msleep %>% 
  group_by(vore) %>% 
  summarise_if(is.numeric, mean, na.rm=TRUE)
```

One of the downsides of the aggregate summarise functions is that you do not require a new column title. It therefore might not always be clear what this new value is (average? median? minimum?). Luckily thanks to similar `rename_*()` functions, it only takes one line extra to rename them all:

```{r}
msleep %>% 
  group_by(vore) %>% 
  summarise_if(is.numeric, mean, na.rm=TRUE) %>% 
  rename_if(is.numeric, ~paste0("avg_", .))
```


<br>

### **Summarise_at**

The function `summarise_at()` also requires two arguments: 

+ First it needs information about the columns you want it to consider. In this case you need to wrap them inside a `vars()` statement. Inside `vars()` you can use anything that can be used inside a `select()` statement. Have a look here if you need more info. 

+ Secondly, it needs information about how to summarise that data, which as above needs to be a function. If not a function, you can create a function on the fly using `funs()` or a tilde (see above).


The sample code below will return the average of all columns which contain the word 'sleep', and also rename them to "avg_*var*" for clarity.

```{r}
msleep %>%
  group_by(vore) %>% 
  summarise_at(vars(contains("sleep")), mean, na.rm=TRUE) %>% 
  rename_at(vars(contains("sleep")), ~paste0("avg_", .))
```




<br><hr>


## **Arranging rows**


```{r}

```


It's useful if your summary tables are arranged, which is `arrange()`'s job. The default format for numeric variables is to sort ascending, but you can add the `desc()` function in your call to change the default. For string variables, it will sort alphabetically.

Sorting numeric variables:  
`arrange(sleep_total)` will arrange it from short sleepers to long sleepers. In this case I wanted the opposite:

```{r}
msleep %>% 
  group_by(vore) %>% 
  summarise(avg_sleep = mean(sleep_total)) %>% 
  arrange(desc(avg_sleep))
```


If you already grouped your data, you can refer to that group within the `arrange()` statement as well by adding a `.by_group = TRUE` statement. This will sort by descending total sleep time, but within each group. 

```{r}
msleep %>% 
  select(order, name, sleep_total) %>% 
  group_by(order) %>% 
  arrange(desc(sleep_total), .by_group = TRUE)
```




<br><hr>


## **Showing only part of your data**

In some cases, you don't just want to show all rows available. Here are some nice shortcuts which can save time


### The 5 lowest and highest values

In some cases, you don't just want to show all rows available. You can filter of course, but there are some shortcuts for specific needs: if you want to select the highests 5 cases, you could combine an `arrange` call with a `head(n=5)`. Or you can use `top_n(5)` which will retain (unsorted) the 5 highest values.

```{r}
msleep %>% 
  group_by(order) %>% 
  summarise(average = mean(sleep_total)) %>% 
  top_n(5)
```


The five lowest values can be found using `top_n(-5)`:

```{r}
msleep %>% 
  group_by(order) %>% 
  summarise(average = mean(sleep_total)) %>% 
  top_n(-5)
```

If you have more than one column, you can add the variable you want it to use. The sample code will retain the 5 highest values of average_sleep.


```{r}
msleep %>% 
  group_by(order) %>% 
  summarise(average_sleep = mean(sleep_total), max_sleep = max(sleep_total)) %>% 
  top_n(5, average_sleep)
```


### A random selection of rows

Using `sample_n()` you can sample a random selection of rows.  
Alternative is `sample_frac()` allowing you to randomly select a fraction of rows (here 10%).


```{r}
msleep %>% 
  sample_frac(.1)
```



### A user-defined slice of rows

The `head()` call will standard show the first 6 rows, which can be modified by adding a n-argument: `head(n=10)`. 
Similarly `tail()` will show the final 6 rows, which again can be modified by adding a n-argument.
If you want to slice somewhere in the middle, you can use `slice()`. The sample code will show rows 50 to 55.


```{r}
msleep %>% 
  slice(50:55)
```


