---
title: "Dplyr03_Filtering_rows"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, error = FALSE, warning=FALSE)
```


# Data Wrangling Part 3: Basic and more advanced ways to filter the cases you need

This is the third blog post in a series of dplyr tutorials:

* [Part 1: Basic to Advanced Ways to Select Columns](https://suzanbaert.netlify.com/2018/01/dplyr-tutorial-1/)
* [Part 2: Transforming your columns into the right shape](https://suzan.rbind.io/2018/02/dplyr-tutorial-2/)


Content:  

--


**The data**  
As per previous blog posts, many of these functions truly shine when you have a lot of columns, but to make it easy on people to copy paste code and experiment, I'm using a built-in dataset:
```{r}
library(tidyverse)

glimpse(msleep)
```
<br>



## **Basic row filters**

In many cases you don't want to include all rows in your analysis but only a selection of rows. The function to use only specific rows is called `filter()` in dplyr. 

### **Filtering rows based on a numeric variable**

You can filter numeric variables based on their values. The most used operators for this are `>`, `>=`, `<`, `<=` and `==`.  

```{r}
msleep %>% 
  select(name, sleep_total) %>% 
  filter(sleep_total > 18)
```

If you want to select a range of values you can use two logical requirements. For instance to select all animals with a total sleep time between 15 and 18 hours, I could use: `filter(sleep_total >= 15, sleep_total <= 18)`, but there is a slightly shorter way by using the `between()` function.

```{r}
msleep %>% 
  select(name, sleep_total) %>% 
  filter(between(sleep_total, 15, 18))
```
<br>

### **Filtering based on a exact character variable matches**

If you want to select a specific group of animals  for instance you can use the `==` comparison operator:


```{r}
msleep %>% 
  select(order, name, sleep_total) %>% 
  filter(order == "Didelphimorphia")
```

Simarly you can use the other operators:  
`filter(order != "Rodentia")` will select everything except the Rodentia rows.  
`filter(name > "v")` will just select the rows with a name in the alphabet after the letter v.  
  
If you want to select more than one animal you can use the `%in%` operator. The following code will just select the rows with animals belonging to the order of Didelphimorphia and Diprotodontia.  

```{r}
msleep %>% 
  select(order, name, sleep_total) %>% 
  filter(order %in% c("Didelphimorphia", "Diprotodontia"))
```

You can use the `%in%` operator to deselect certain groups as well, in this case you have to negate by adding an exclamation mark at the beginning of your `filter`. Making a `!%in%` might seem logic but it won't work.

```{r}
remove <- c("Rodentia", "Carnivora", "Primates")
msleep %>% 
  select(order, name, sleep_total) %>% 
  filter(!order %in% remove)
  
```

<br>

### **Filtering rows based on regex**

The above options will only work if you can use the full variable content. In some cases though it will be needed to filter based on partial matches. In this case, we need a function that will evaluate regular expressions on strings and return boolean values. Whenever the statement is `TRUE` the row will be filtered.  
There are two main options for this: base R's `grepl()` function, or `str_detect()` from the `stringr` package. 

Whenever you are lookin for partial matches, it is important to remember that R is case sensitive. By just using `filter(str_detect(name, pattern="mouse"))` we would leave out any row called Mouse. In this case it does not make a difference, but it's a good habit to create. 

I used `str_detect()` below as it is easier to understand for beginners. For those intersted, the alternative would be `filter(grepl(pattern="mouse", tolower(name)))`.  

```{r}
msleep %>% 
  select(name, sleep_total) %>% 
  filter(str_detect(tolower(name), pattern = "mouse"))

  
```

<br>

### **Filtering based on multiple conditions**

All the above just returns rows based on a single condition, but the filter option allows also AND and OR style filters:  
`filter(condition1 | condition2)` will return rows where condition 1 OR condition 2 is met.   
`filter(condition1, condition2)` will return rows where both conditions are met.  
  
Multipe AND, OR and NOT conditions can be combined. The sample code will return all rows with a bodywt above 100 and either have a sleep_total above 15 or are not part of the Carnivora order.

```{r}
msleep %>% 
  select(name, order, sleep_total:bodywt) %>% 
  filter(bodywt > 100, (sleep_total > 15 | order != "Carnivora"))
  
```

<br>

## **Filtering across multiple columns**

### **Filter_all**

Admittedly, `msleep` is not the best database to showcase this power, but imagine you have a database with a few columns and you want to select all rows that have a certain word in either column. Take a financial dataframe for instance and you want to select all rows with 'food', whether food is mentioned in the main category column, the subcategory column, the comments column or the place you've spent it.  
You could make a long filter statement with 4 different conditions wrapped inside OR statements. Or you just filter across all columns for the string "food".  

In the sample code below I'm searching for the string "Ca" across all columns.  
The equivalent of 'OR' in this case is `any_vars()`. The below code basically asks to retain any rows where any of the variables has the pattern "Ca" inside.


```{r}
msleep %>% 
  filter_all(any_vars(str_detect(., pattern = "Ca")))
```


The same can be done for numerical values:
This code will retain any rows that has any value below 0.1:


```{r}
msleep %>%  
  select(name, sleep_total:bodywt) %>% 
  filter_all(any_vars(. < 0.1))
  
```

The `any_vars()` statement is equivalent to OR, so of course there is an equivalent for AND statements as well:`all_vars()`.
The below code will retain any rows where all values are above 1.

```{r}
msleep %>%  
  select(name, sleep_total:bodywt, -awake) %>% 
  filter_all(all_vars(. > 1))
  
```

<br>

### **Filter_if**

The `filter_all() can sometimes go a bit wild. In the msleep database I have a set of sleep and weight measurements where some data is missing - there is nothing I can do to add data there. But the first few set of columns just contain info on animals. The vore of Vesper Mouse is missing, but that is info I can still dig up and add to the dataframe if I wanted.  
So imagine I want to find out all data rows where we NA in the first few columns. `filter_all(any_vars(is.na(.)))` will be quite useless because it would return 27 rows, many of which are missing data in the measurement section.  

In this case: `filter_if()` comes in handy. The describing columns are all character columns, while the measurement data is numeric. So using `filter_if()` I can specify that I want to just filter on character variables. In this case I onnly get 7 rows.


```{r}
msleep %>% 
  select(name:order, sleep_total:sleep_rem) %>% 
  filter_if(is.character, any_vars(is.na(.)))

  
```

Similarly, you can add `is.numeric`, `is.integer`, `is.double`, `is.logical`, `is.factor`.
If you have data columns, you can load the lubridate package, and use `is.POSIXt` or `is.Date`.

<br>



### **Filter_at**

One of the more powerful functions is `filter_at()`: it does not filter all columns, nor does it need you to specify the type of column, you can just select columns to which the change should happen via the `vars()` argument. This argument allows anything that can be done within a select statement: so you can use logical numerical functions, regex, etc (See my first blog post for select options).  
The second argument is the condition for selection. You can use `all_vars()` if all columns need to return TRUE (AND equivalent), or `any_vars()` in case just one variable needs to return TRUE (OR equivalent).


```{r}
msleep %>% 
  select(name, sleep_total:sleep_rem, brainwt:bodywt) %>% 
  filter_at(vars(contains("sleep")), all_vars(.>5))
  
```
