---
title: "blogpost2"
output:
  md_document:
    toc: true
    toc_depth: 3
---

This is a second post in a series of dplyr functions.
The first one covered basic and advanced ways to select, rename and reorder columns,and can be found here.   
  
This second one covers tools to manipulate your columns to get them the way you want them. 
- content



**The data**  
As per previous blog posts, many of these functions truly shine when you have a lot of columns, but to make it easy on people to copy paste code and experiment, I'm using a built-in dataset:
```{r message=FALSE, warning=FALSE}
library(tidyverse)

glimpse(msleep)
```
<br>


## **Mutating columns: the basics**

You can make new columns with the `mutate()` function. The options inside mutate are almost endless: pretty much anything that you can do to normal vectors, can be done inside a `mutate()` function.  
Anything inside `mutate` can either be a new column (by giving mutate a new column name), or can replace the current column (by keeping the same column name).  
  
One of the simplest options is a calculation based on values in other columns. In the sample code, we're changing the sleep data from data measured in hours to minutes. 

```{r}
msleep %>%
  select(name, sleep_total) %>%
  mutate(sleep_total_min = sleep_total * 60)
```


New columns can be made with aggregate functions such as average, median, max, min, sd, ...  
The sample code makes two new columns: one showing the difference of each row versus the average sleep time, and one showing the difference versus the animal with the least sleep.

```{r}
msleep %>%
  select(name, sleep_total) %>%
  mutate(sleep_total_vs_AVG = sleep_total - round(mean(sleep_total), 1),
         sleep_total_vs_MIN = sleep_total - min(sleep_total))
```


In the below comments, Steve asked about aggregate functions across columns. These functions by nature will want to summarise a column (like shown above), if however you want to `sum()` or `mean()` across columns, you might run into errors or absurd answers. In these cases you either can revert to actually spelling out the arithmetics: `mutate(average = (sleep_rem + sleep_cycle) / 2)` or you have to add a special instruction to the pipe that it should perform these aggregate functions not on the entire column, but by row: 

```{r}
#alternative to using the actual arithmetics:
msleep %>% 
  select(name, contains("sleep")) %>% 
  rowwise() %>% 
  mutate(avg = mean(c(sleep_rem, sleep_cycle)))
```



The `ifelse()` function deserves a special mention because it is particularly useful if you don't want to mutate the whole column in the same way. With `ifelse()`, you first specify a logical statement, afterwards what needs to happen if the statement returns `TRUE`, and lastly what needs to happen if it's  `FALSE`.  
  
Imagine that we have a database with two large values which we assume are typos or measurement errors, and we want to exclude them. The below code will take any `brainwt` value above 4 and return NA. In this case, the code won't change for anything below 4.  

```{r}
msleep %>%
  select(name, brainwt) %>%
  mutate(brainwt2 = ifelse(brainwt > 4, NA, brainwt)) %>%
  arrange(desc(brainwt))
```


You can also mutate string columns with stringr's `str_extract()` function in combination with any character or regex patterns.  
The sample code will return the last word of the animal name and makes it lower case.

```{r}

msleep %>%
  select(name) %>%
  mutate(name_last_word = tolower(str_extract(name, pattern = "\\w+$")))
```

<br>



## **Mutating several columns at once**

This is where the magic really happens. Just like with the `select()` functions in part 1, there are variants to `mutate()`: 

+ `mutate_all()` will mutate all columns based on your further instructions
+ `mutate_if()` first requires a function that returns a boolean to select columns. If that is true, the mutate instructions will be followed on those variables.
+ `mutate_at()` requires you to specify columns inside a `vars()` argument for which the mutation will be done.



### **Mutate all**

The `mutate_all()` version is the easiest to understand, and pretty nifty when cleaning your data. You just pass an action (in the form of a function) that you want to apply across all columns.

Something easy to start with: turning all the data to lower case:


```{r}
msleep %>%
  mutate_all(tolower)
```


The mutating action needs to be a function: in many cases you can pass
the function name without the brackets, but in some cases you need
arguments or you want to combine elements. In this case you have some
options: either you make a function up front (useful if it's longer), or
you make a function on the fly by wrapping it inside `funs()` or via a
tilde.

The below paste mutation requires a function on the fly. You can
either use `~paste(., "  /n  ")` or
`funs(paste(., "  /n  "))`. When making a function on the
fly, you usually need a way to refer to the value you are replacing: which is what the `.` symbolizes.

For instance, after scraping the web, you often have tables with too many spaces and extra `\n` signs, but you can clean it all in one go.

I'm first going to use `mutate_all()` to screw things up:
```{r}
msleep_ohno <- msleep %>%
  mutate_all(~paste(., "  /n  "))

msleep_ohno[,1:4]
```

Let's clean it up again:   
In this code I am assume that not all values show the same amount of
extra white spaces as is often the case with parsed data: it first
removes any `/n`, and then trims any additional white spaces:
```{r}
msleep_corr <- msleep_ohno %>%
  mutate_all(~str_replace_all(., "/n", "")) %>%
  mutate_all(str_trim)

msleep_corr[,1:4]
```




### **Mutate if**

Not all cleaning functions can be done with `mutate_all()`. Trying to round your data will lead to an error if you have both numerical and character columns. 

```{r eval = FALSE}
msleep %>%
  mutate_all(round)
```

`Error in mutate_impl(.data, dots) : Evaluation error: non-numeric argument to mathematical function.` 


In these cases we have to add the condition that columns need to be
numeric before giving `round()` instructions, which can be done with mutate_if.

By using `mutate_if()` we need two arguments inside a pipe:

+ First it needs information about the columns you want it to consider. This information needs to be a function that returns a boolean value. The easiest cases are functions like `is.numeric`, `is.integer`, `is.double`, `is.logical`, `is.factor`, `lubridate::is.POSIXt` or `lubridate::is.Date`.

+ Secondly, it needs instructions about the mutation in the form of a function. If needed, use a tilde or `funs()` before (see above).


```{r}
msleep %>%
  select(name, sleep_total:bodywt) %>%
  mutate_if(is.numeric, round)
```



###  **Mutate at to change specific columns**

By using `mutate_at()` we need two arguments inside a pipe:

+ First it needs information about the columns you want it to consider. In this case you can wrap any selection of columns (using all the options possible inside a `select()` function) and wrap it inside `vars()`.

+ Secondly, it needs instructions about the mutation in the form of a function. If needed, use a tilde or `funs()` before (see above).


All sleep-measuring columns are in hours. If I want those in minutes, I can use `mutate_at()` and wrap all 'sleep' containing columns inside `vars()`. Secondly, I make a function in the fly to multiple every value by 60.  
The sample code shows that in this case all `sleep` columns have been changed into minutes, but `awake` did not.
```{r}
msleep %>%
  select(name, sleep_total:awake) %>%
  mutate_at(vars(contains("sleep")), ~(.*60)) 
```

###  **Changing multiple columns names**

With a singular `mutate()` statement, you immediately have the option to change the columns name. In the above example for instance it is confusing that the sleep columns are in a different unit, you can change that by calling a rename function:

```{r}
msleep %>%
  select(name, sleep_total:awake) %>%
  mutate_at(vars(contains("sleep")), ~(.*60)) %>% 
  rename_at(vars(contains("sleep")), ~paste0(.,"_min"))
```


Or as [Tomas
McManus](https://twitter.com/TomasMcManus1/status/981187099649912832)
pointed out: you can assign a "tag" inside `funs()` which will be appended to the current name. The
main difference between both options: the `funs()` version is one line
of code less, but columns will be added rather than replaced. Depending
on your scenario, either could be useful.

```{r}
msleep %>%
  select(name, sleep_total:awake) %>%
  mutate_at(vars(contains("sleep")), funs(min = .*60))
```


<br><hr>


## **Working with discrete columns**

### **Recoding discrete columns**

To rename or reorganize current discrete columns, you can use `recode()` inside a `mutate()` statement: this enables you to change the current naming, or to group current levels into less levels. 
The `.default` refers to anything that isn't covered by the before groups with the exception of NA. You can change NA into something other than NA by adding a `.missing` argument if you want (see next sample code).

```{r}
msleep %>%
  mutate(conservation2 = recode(conservation,
                        "en" = "Endangered",
                        "lc" = "Least_Concern",
                        "domesticated" = "Least_Concern",
                        .default = "other")) %>%
  count(conservation2)
  
```

A special version exists to return a factor: `recode_factor()`. By default the `.ordered` argument is `FALSE`. To return an ordered factor set the argument to `TRUE`:

```{r}
msleep %>%
  mutate(conservation2 = recode_factor(conservation,
                        "en" = "Endangered",
                        "lc" = "Least_Concern",
                        "domesticated" = "Least_Concern",
                        .default = "other",
                        .missing = "no data",
                        .ordered = TRUE)) %>%
  count(conservation2)
```
<br>

### **Creating new discrete column (two levels)**

The `ifelse()` statement can be used to turn a numeric column into a discrete one. As mentioned above, `ifelse()` takes a logical expression, then what to do if the expression returns `TRUE` and lastly what to do when it returns `FALSE`.   
The sample code will divide the current measure `sleep_total` into a discrete "long" or "short" sleeper.

```{r}
msleep %>%
  select(name, sleep_total) %>%
  mutate(sleep_time = ifelse(sleep_total > 10, "long", "short")) 
```
<br>

### **Creating new discrete column (multiple levels)**

The `ifelse()` can be nested but if you want more than two levels, but it might be even easier to use `case_when()` which allows as many statements as you like and is easier to read than many nested `ifelse` statements.    
The arguments are evaluated in order, so only the rows where the first statement is not true will continue to be evaluated for the next statement. For everything that is left at the end just use the `TRUE ~ "newname"`.  
Unfortunately there seems to be no easy way to get `case_when()` to return an ordered factor, so you will need to to do that yourself afterwards, either by using `forcats::fct_relevel()`, or just with a `factor()` function. If you have a lot of levels I would advice to make a levels vector upfront to avoid cluttering the piple too much.

```{r}
msleep %>%
  select(name, sleep_total) %>%
  mutate(sleep_total_discr = case_when(
    sleep_total > 13 ~ "very long",
    sleep_total > 10 ~ "long",
    sleep_total > 7 ~ "limited",
    TRUE ~ "short")) %>%
  mutate(sleep_total_discr = factor(sleep_total_discr, 
                                    levels = c("short", "limited", 
                                               "long", "very long")))
```



The `case_when()` function does not only work inside a column, but can be used for grouping across columns:

```{r}
msleep %>%
  mutate(silly_groups = case_when(
    brainwt < 0.001 ~ "light_headed",
    sleep_total > 10 ~ "lazy_sleeper",
    is.na(sleep_rem) ~ "absent_rem",
    TRUE ~ "other")) %>%
  count(silly_groups)
```

<br><hr>




## **Splitting and merging columns**

Take for example this dataset
```{r message=FALSE, warning=FALSE}
(conservation_expl <- read_csv("conservation_explanation.csv"))
```


You can unmerge any columns by using tidyr's `separate()` function. To do this, you have to specify the column to be splitted, followed by the new column names, and which seperator it has to look for.  
The sample code shows seperating into two columns based on '=' as a separator.

```{r}
(conservation_table <- conservation_expl %>%
  separate(`conservation abbreviation`, 
           into = c("abbreviation", "description"), sep = " = "))
```


The opposite is tidyr's `unite()` function. You specify the new column name, and then the columns to be united, and lastly what seperator you want to use.  


```{r}
conservation_table %>%
  unite(united_col, abbreviation, description, sep=": ")
```

<br>

## **Bringing in columns from other data tables**

If you want to add information from another table, you can use the joining functions from `dplyr`. The msleep data contains abbreviations for conservation but if you are not familiar with the topic you might need the description we used in the section above inside the msleep data.  
  
Joins would be a chapter in itself, but in this particular case you would do a `left_join()`, i.e. keeping my main table (on the left), and adding columns from another one to the right. In the `by = ` statement you specify which colums are the same, so the join knows what to add where.  
The sample code will add the description of the different conservation states into our main `msleep` table. The main data contained an extra `domisticated` label which i wanted to keep. This is done in the last line of the table with an `ifelse()`.


```{r}
msleep %>%
  select(name, conservation) %>%
  mutate(conservation = toupper(conservation)) %>%
  left_join(conservation_table, by = c("conservation" = "abbreviation")) %>%
  mutate(description = ifelse(is.na(description), conservation, description))
```
<br>


## **Spreading and gathering data**

The `gather()` function will gather up many columns into one. In this case, we have 3 columns that describe a time measure. For some analysis and graphs, it might be necessary to get them all into one.  
The `gather` function needs you to give a name ("key") for the new descriptive column, and a another name ("value") for the value column. The columns that you don't want to gather need to be deselected at the end. In the sample code I'm deselecting the column `name`.  

```{r}
msleep %>%
  select(name, contains("sleep")) %>%
  gather(key = "sleep_measure", value = "time", -name)
```

A useful attribute in gathering is the `factor_key` argument which is `FALSE`by default. In the previous example the new column `sleep_measure` is a character vector. If you are going to summarise or plot afterwards, that column will be ordered alphabetically.  
If you want to preserve the original order, add `factor_key = TRUE` which will make the new column an ordered factor.


```{r}
(msleep_g <- msleep %>%
  select(name, contains("sleep")) %>%
  gather(key = "sleep_measure", value = "time", -name, factor_key = TRUE))
```


The opposite of gathering is spreading. Spread will take one column and make multiple columns out of it. If you would have started with the previous column, you could get the differrent sleep measures in different columns:

```{r}
msleep_g %>%
  spread(sleep_measure, time)
```



## **Turning data into NA**

The function `na_if()` turns particular values into `NA`. In most cases the command probably be `na_if("")` (i.e turn an empty string into NA), but in principle you can do anything.

The same code will turn any value that reads "omni" into NA

```{r}
msleep %>%
  select(name:order) %>%
  na_if("omni")
```

