---
title: "JSON_parsing"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(jsonlite)
```

## Introduction

There are plenty of tutorials on taking JSON strings and returning R objects. I guess this is what most data analysts/scientists are doing: grabbing data from an API and analyzing in R.  

For a recent project, I wanted to do the opposite: **use R to upload my data to an API**. The amount of tutorials I found on that topic can be counted on one hand, and most of them did not cover more complex nested structures. So instead of learning from tutorials, I started experimenting quite a lot. I do not know whether there are any best practices on parsing to JSON from R: I think it all depends on the structure of the data you have, and the final structure required for JSON parsing. I most certainly do not have the expertise for saying what the best practices are, but I can share some of the tricks I found along the way that might aid the next person wandering into this territory.




## What are JSON files?

- why
- Syntax of JSON


### Basic JSON parsing

There will be plenty of examples belowm, but in general it boils down to this:

+ You obtain a JSON object by parsing a list
+ You obtain a JSON array by parsing a dataframe (or by a list of a list)
+ By nesting dataframes/lists inside lists/dataframes you can get the required nested JSON structures. In many cases you can get there starting from a list or a dataframe, so you can pick the easiest option depending on the shape of your data to start with.



A standard list will return a JSON object:


```{r}
r_object <- list(key1 = "value1", 
                 key2 = "value2")

toJSON(r_object, auto_unbox = TRUE, pretty = TRUE)
```



A standard dataframe will return a JSON array:

```{r}
r_object <- data.frame(key1 = "value1", 
                       key2 = "value2")

toJSON(r_object, pretty = TRUE)
```

The same JSON array can be obtained by wrapping your list inside a lists:

```{r}
r_object <- list(
                list(key1 = "value1",
                     key2 = "value2"))
toJSON(r_object, auto_unbox = TRUE, pretty = TRUE)
```




### Viewing JSON in R

```{r include= FALSE}

json_example_list <- list(type = "tutorial", topic = "json parsing", packages_used = c("jsonlite", "dplyr", "purrr", "tidyr"))
json_example <- jsonlite::toJSON(json_example_list, auto_unbox = TRUE)
```


This is a json_example in its non-intended form which is not easily human readable.
```{r}
json_example
```

To show it in its more-readable indented form you can use `jsonlite::prettify()`

```{r}
prettify(json_example)
```

Similarly, when parsing an R object to JSON with the `jsonlite::toJSON()` function, by default you get the non-indented form, but by adding the argument `pretty = TRUE` (as will be done throughout this tutorial) it will print its indented form. 




### Note about sending to an API

Add here that by itself it will already be parsed in the post function. however to control its json structure, it's good to experiment parsing upfront already. If you want to send already parsed data you have to add that one argument.

learning in POST requests
then adding no json parsing


<br><hr>



## Tutorial set-up

Packages we will need: dplyr for data wrangling, and jsonlite to parse data to JSON.

```{r}
library(jsonlite)
library(dplyr)
```


To make it easy, I want to use a dataset most people will have available on their pc. We're going to pretend that we have a database with animals and their sleep time, and we recently gathered two more data points (a cat and a rabbit) that we want to upload to the database. I also made a selection of columns (6 rather than all 11) to save some space.

```{r}
msleep <- ggplot2::msleep
cat_rabbit <- msleep$name %in% c("Rabbit", "Domestic cat")
columns <- c(1:3, 6, 10:11)
new_data <- msleep[cat_rabbit, columns]
new_data
```






<br><hr>

## Basic unnested JSON structures

You might get lucky: the easiest case is that you can bulk add all your observations, and that the required json string is an unnested array of objects. In this case, you can just parse all your data to JSON and you're done.

```{r}
toJSON(new_data, pretty = TRUE)
```


If bulk uploading is not available and you need to submit row by row, it depends whether the api will ask for a JSON array or JSON object. If it's a JSON array, you can send the data row by row, and build a function around the below JSON call.

```{r}
toJSON(new_data[1, ], pretty = TRUE)
```


If it needs to be JSON object, you need to first get to a list. If you do not have many elements you could simply build a list manually and refer to each element in the datatable, but with many columns that's a lot of manual labour.  

There are two others ways: 
1. Iterate over every row of your data and call `as.list` on that row:  


```{r}
toJSON(as.list(new_data[1, ]), auto_unbox = TRUE, pretty = TRUE)
```

or 2. First tranpose your entire datatable to an inside-out-list, and send each element of the list seperately. 
In both of these case I also added the `auto_unbox = TRUE` argument to overwrite toJSONs default behaviour to turn list elements of 1 into an array of length 1.

```{r}
transposed_data <- purrr::transpose(new_data)
toJSON(transposed_data[[1]], auto_unbox = TRUE, pretty = TRUE)
```


The difference between a straight `as.list()` call and an inside-out `transpose()` list call can best be shown with an example. Below you can see that `as.list()` will structure the list by column, while `transpose()` lists by row. As you can imagine, both cases would result in very different JSON structures.

```{r}
as.list(new_data)
```


```{r}
purrr::transpose(new_data)
```



<br><hr>


## Nested JSON structures

When it comes to nested structures, it depends on the api structure and the data you already have. In some cases you might want to start from your data frame and embed it into a list, or nest parts of it. 



### Embed your data inside a list

Imagine your dataframe already covers the nested part of the required json string, but it needs some additional info one level up. One way is to embed your data inside a list. You can embed one or more dataframes, and those can be identified upfront or assembled on the spot.


```{r}
r_object <- list(measurement_date = Sys.Date(),
                 measurement_place = "Belgium",
                 data = new_data)

toJSON(r_object, auto_unbox = TRUE, pretty = TRUE)
```


Below is an example of two embedded dataframes, one of which is made at the time of the list call. Both dataframes will be converted to nested JSON arrays.

```{r}
r_object <- list(submission_date = Sys.Date(),
                 measurement_info = data.frame(date = "2018-04-09", place = "Belgium"),
                 data = new_data)

toJSON(r_object, auto_unbox = TRUE, pretty = TRUE)
```


If you need one nested JSON object instead of an array, you need to turn it into a list. You can do this on the spot (as shown below), or turn the data into the right list upfront using `as.list()` or `transpose()` as shown above.


```{r}
r_object <- list(submission_date = Sys.Date(),
                 measurement_info = list(date = "2018-04-09", place = "Belgium"),
                 data = new_data)

toJSON(r_object, auto_unbox = TRUE, pretty = TRUE)

```



<br><hr>

### Nested dataframes





---

### Parsing unnest JSON object

When you start from an unnested list, you will parse into a JSON object:


If every element in your list has length 1, the `toJSON()` default is to place that one element in an array, which is very often what you don't want to do. In this case, the solution is to add the auto_unbox = TRUE argument, which results in a more standard `"key":"Value"` pair.

```{r}
from_list <- list(
  country_name = "Belgium",
  capital = "Brussels",
  population = 11358357)

toJSON(from_list, auto_unbox = TRUE, pretty = TRUE)
```

If the length of your list elements is more than 1, you end up with a JSON object that contains an array for every key.

```{r}
from_list <- list(
  country_name = c("Belgium", "France"),
  capital = c("Brussels", "Paris"),
  population = c(11358357, 67201000))

toJSON(from_list, pretty = TRUE)
```


### Parsing basic (unnested) JSON arrays

What if you need a JSON array rather than an object? 
By making an unnested dataframe rather than a list, you will get a JSON array with 1 element.

```{r}
from_df <- data.frame(
  country_name = "Belgium",
  capital = "Brussels",
  population = 11358357)

toJSON(from_df, pretty = TRUE)
```


If your dataframe contains more than one row, every row will be an element of your JSON array:

```{r}
from_df <- data.frame(
  country_name = c("Belgium", "France"),
  capital = c("Brussels", "Paris"),
  population = c(11358357, 67201000))

toJSON(from_df, pretty = TRUE)
```

<br><hr>

In many cases the JSON structures you need are not unnest one-dimentional JSON objects or arrays. 


```{r}
(from_df <- data.frame(
  country_name = c("Belgium", "Belgium", "Belgium", "France"),
  language = c("Dutch", "French", "German", "French"),
  language_population = c("56%", "38%", "1%", "100%"), stringsAsFactors = FALSE))
```

```{r}
(df_nest <- tidyr::nest(from_df, -country_name, .key = "official_languages"))
```



```{r}
toJSON(df_nest, pretty = TRUE)
```
```



## Parsing nested structures

To save on space, I made a dataframe `iris_small` which only contains 2 rows per Species.


```{r}
iris_small <- iris %>%
  group_by(Species) %>%
  top_n(2, Sepal.Width)

toJSON(iris_small, pretty = TRUE)


```


```{r}
iris_nest <- iris_small %>%
  tidyr::nest(-Species, .key = "measurements") 

toJSON(iris_nest, pretty = TRUE)
```







#### Data sources

https://nl.wikipedia.org/wiki/Talen_in_Belgi%C3%AB
