---
title: "JSON_parsing"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

There are plenty of tutorials on taking JSON strings and returning R objects. I guess this is what most data scientists are doing: grabbing data from an API and analyzing in R.  

For this project, I was doing the opposite: People are submitting their data, R is doing some pre-analysis wrangling and is sending it to an API.



## What are JSON files?

- why
- Syntax of JSON

## JSON in R

```{r}
library(jsonlite)
library(dplyr)
library(tidyr)
```


First explain prettify/minify and toJSON with pretty=TRUE


## Basic JSON structures

### Parsing basic (unnested) JSON objects

When you start from an unnested list, you will parse into a JSON object:

```{r}
from_list <- list(
  country_name = "Belgium",
  capital = "Brussels",
  population = 11358357)

toJSON(from_list, pretty = TRUE)
```

If every element in your list has length 1, the `toJSON()` default is to place that one element in an array, which is very often what you don't want to do. In this case, the solution is to add the auto_unbox = TRUE argument, which results in a more standard `"key":"Value"` pair.

```{r}
from_list <- list(
  country_name = "Belgium",
  capital = "Brussels",
  population = 11358357)

toJSON(from_list, auto_unbox = TRUE, pretty = TRUE)
```

If the length of your list elements is more than 1, you end up with a JSON object that contains an array for every key.

```{r}
from_list <- list(
  country_name = c("Belgium", "France"),
  capital = c("Brussels", "Paris"),
  population = c(11358357, 67201000))

toJSON(from_list, pretty = TRUE)
```


### Parsing basic (unnested) JSON arrays

What if you need a JSON array rather than an object? 
By making an unnested dataframe rather than a list, you will get a JSON array with 1 element.

```{r}
from_df <- data.frame(
  country_name = "Belgium",
  capital = "Brussels",
  population = 11358357)

toJSON(from_df, pretty = TRUE)
```


If your dataframe contains more than one row, every row will be an element of your JSON array:

```{r}
from_df <- data.frame(
  country_name = c("Belgium", "France"),
  capital = c("Brussels", "Paris"),
  population = c(11358357, 67201000))

toJSON(from_df, pretty = TRUE)
```

<br><hr>

In many cases the JSON structures you need are not unnest one-dimentional JSON objects or arrays. 


```{r}
(from_df <- data.frame(
  country_name = c("Belgium", "Belgium", "Belgium", "France"),
  language = c("Dutch", "French", "German", "French"),
  language_population = c("56%", "38%", "1%", "100%"), stringsAsFactors = FALSE))
```

```{r}
(df_nest <- nest(from_df, -country_name, .key = "official_languages"))
```



```{r}
toJSON(df_nest, pretty = TRUE)
```
```



## Parsing nested structures

To save on space, I made a dataframe `iris_small` which only contains 2 rows per Species.


```{r}
iris_small <- iris %>%
  group_by(Species) %>%
  top_n(2, Sepal.Width)

toJSON(iris_small, pretty = TRUE)


```


```{r}
iris_nest <- iris_small %>%
  tidyr::nest(-Species, .key = "measurements") 

toJSON(iris_nest, pretty = TRUE)
```







#### Data sources

https://nl.wikipedia.org/wiki/Talen_in_Belgi%C3%AB
